---
title: "Analysis Notebook"
description: "Analysis notebook for NCE kōura catfish experiment"
format:
  html:
    fig-path: "analysis_files/figure-html/"
    theme:
      light: flatly
      dark: darkly
    code-fold: true
    code-tools: true
    code-summary: "View Code"
    toc: true
    toc-depth: 3
    number-sections: true
fig-width: 8
fig-height: 6

execute:
  freeze: false
  cache: false

---

# Setup
```{r}
#| label: setup
#| include: false

packages <- c("ggpattern","multcomp","DT","car","glmmTMB","AICcmodavg", "ggeffects","DHARMa", "brms", "emmeans", "lme4","lmerTest","vegan","ggnewscale","multcompView","patchwork","janitor","lubridate","stringr","tidyverse","dplyr","ggplot2","readxl","writexl","readr")

# Load packages if not already installed
quiet_load <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    suppressWarnings(suppressMessages(install.packages(pkg, dependencies = TRUE)))}
  suppressPackageStartupMessages(require(pkg, character.only = TRUE, quietly = TRUE))
  invisible(TRUE)}

invisible(lapply(packages, quiet_load))

exc_file_dir <- "data/raw/NCE_data.xlsx"
der_data_dir <- "data/derived"
out_dir      <- "outputs"

# Read sheets from raw xlsx
df_NCE_raw <- readxl::read_excel(exc_file_dir, sheet = "NCE_exp")
df_Feeding_raw <- readxl::read_excel(exc_file_dir, sheet = "Feeding")

# Save as csv to derived/
write.csv(df_NCE_raw,     file.path(der_data_dir, "NCE_exp.csv"),  row.names = FALSE)
write.csv(df_Feeding_raw, file.path(der_data_dir, "Feeding.csv"),  row.names = FALSE)

df_NCE_raw     <- read.csv(file.path(der_data_dir, "NCE_exp.csv"))
df_Feeding_raw <- read.csv(file.path(der_data_dir, "Feeding.csv"))

# Set base for plots
base_theme_bw <- theme_classic() +
  theme(text     = element_text(family = "sans", size = 11),
    axis.title   = element_text(face = "plain"),
    axis.text    = element_text(face = "plain"),
    plot.title   = element_text(face = "plain"),
    strip.text   = element_text(face = "plain"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.3))
theme_set(base_theme_bw)

# Format of p values in text
format_pval <- function(p) {
  if (p < 0.001) {
    return("<0.001")
  } else {
    return(format(round(p, 3), nsmall = 3))
  }
}

# Get treatment contrast
get_treat_contrast <- function(condition, contrast) {
  row <- emmeans_stratified %>% 
    filter(condition == !!condition, contrast == !!contrast)
  
  if(nrow(row) == 0) {
    return("Contrast not found")
  }
  
  p <- format_pval(row$`p.value`[1])
  p
}

# Get period contrast p-value
get_period_contrast <- function(condition, contrast) {
  row <- emmeans_stratified %>% 
    filter(condition == !!condition, contrast == !!contrast)
  format_pval(row$`p.value`)
}

# Get OCL correlation
get_ocl_corr <- function(treatment) {
  if(treatment == "catfish") {
    rs <- round(spearman_catfish$estimate, 2)
    p <- format_pval(spearman_catfish$p.value)
  } else if(treatment == "eel") {
    rs <- round(spearman_eel$estimate, 2)
    p <- format_pval(spearman_eel$p.value)
  } else {
    rs <- round(spearman_control$estimate, 2)
    p <- format_pval(spearman_control$p.value)
  }
  paste0("rs = ", rs, ", p = ", p)
}

# Get light overall effect
get_light_effect <- function() {
  p <- format_pval(summary(m_light_effect)$coefficients$cond["lightdark", "Pr(>|z|)"])
  paste0("p = ", p)
}

```

# Refuge experiment
## Build refuge dataframe
```{r}
#| label: build-df
#| include: false

df_NCE_raw <- df_NCE_raw %>%
  dplyr::select(-location_T180, -behaviour_T180)

# make into long format
koura_long <- df_NCE_raw %>%
  pivot_longer(
    cols = matches("^(location|behaviour)_"),
    names_to   = c(".value","time"),
    names_pattern = "(location|behaviour)_T?(\\d+)" ) %>%
  mutate(time = as.numeric(time))

koura_long <- koura_long %>%
  mutate(period = case_when(
    time < 60 ~ "before",
    time >= 60 & time < 240 ~ "during",
    time >= 240 ~ "after",
    TRUE ~ NA_character_))%>%
  mutate(period = factor(period, levels = c("before", "during", "after")))

#Group all refuge used
koura_long_ref <- koura_long %>%
  mutate(
    location = case_when(
      location %in% c("floor", "wall", "barrier") ~ "open",
      location %in% c("ref_back", "ref_open", "ref_next") ~ "refuge",
      TRUE ~ NA_character_))

koura_long_ref <- koura_long_ref %>%
  mutate(period = case_when(
    time < 60 ~ "before",
    time >= 60 & time < 240 ~ "during",
    time >= 240 ~ "after",
    TRUE ~ NA_character_))%>%
  mutate(period = factor(period, levels = c("before", "during", "after")))

```

## Refuge table
```{r}
#| label: tbl-refuge-prop
#| include: true
#| tbl-cap: "Proportion of kōura observations recorded in refuge across experimental treatments (catfish, eel, control), light conditions (light, dark), and periods (before, during, and after predator exposure). Combined rows summarise across all levels of a given factor. Values represent the proportion of total observations where kōura were located in the refuge."

prop_refuge_light_treat <- koura_long_ref %>%
  dplyr::group_by(light, treatment, period) %>%
  dplyr::summarise(
    n_total  = dplyr::n(),
    n_refuge = sum(location == "refuge", na.rm = TRUE),
    prop_refuge = n_refuge / n_total,.groups = "drop") %>%
  mutate(treatment = factor(treatment, levels = c("catfish","eel","control")),
         light     = factor(light,     levels = c("light", "dark"))) %>%
  arrange(light, treatment, period)

prop_refuge_treat <- koura_long_ref %>%
  dplyr::group_by(treatment, period) %>%
  dplyr::summarise(light = "combined",
    n_total  = dplyr::n(),
    n_refuge = sum(location == "refuge", na.rm = TRUE),
    prop_refuge = n_refuge / n_total,.groups = "drop")

prop_refuge_light <- koura_long_ref %>%
  dplyr::group_by(light, period) %>%
  dplyr::summarise(treatment = "combined",
    n_total  = dplyr::n(),
    n_refuge = sum(location == "refuge", na.rm = TRUE),
    prop_refuge = n_refuge / n_total,.groups = "drop")%>%
  mutate(light     = factor(light,     levels = c("light", "dark"))) %>%
  arrange(light, treatment, period)

prop_refuge_per <- koura_long_ref %>%
  dplyr::group_by(period) %>%
  dplyr::summarise(treatment = "combined",
    light = "combined",
    n_total  = dplyr::n(),
    n_refuge = sum(location == "refuge", na.rm = TRUE),
    prop_refuge = n_refuge / n_total,.groups = "drop")

prop_refuge_comb <- koura_long_ref %>%
  dplyr::summarise(period = "combined",
    treatment = "combined",
    light = "combined",
    n_total  = dplyr::n(),
    n_refuge = sum(location == "refuge", na.rm = TRUE),
    prop_refuge = n_refuge / n_total,.groups = "drop")

refuge_table <- dplyr::bind_rows(prop_refuge_light_treat, prop_refuge_treat,prop_refuge_light, prop_refuge_per, prop_refuge_comb) %>%
  mutate_if(is.numeric, round, digits = 2)

get_prop <- function(l, t, p) {
  refuge_table %>%
    filter(light == l, treatment == t, period == p) %>%
    pull(prop_refuge) %>%
    `*`(100) %>%  # Convert to percentage
    round(1)      # Round to 1 decimal place
}

write.csv(refuge_table, file = file.path(out_dir, "tbl-refuge-prop.csv"), row.names = FALSE)

if (knitr::is_html_output()) {
  datatable(
    refuge_table,
    filter = "top",
    options = list(
      pageLength = 37,
      dom = 'Bfrtip',
      buttons = c('csv', 'excel'),
      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    ),
    extensions = 'Buttons',
    rownames = FALSE,
    colnames = c("Light", "Treatment", "Period", "N total", "N refuge", "Proportion in refuge")
  ) |>
    formatRound(columns = "prop_refuge", digits = 2) |>
    formatStyle(
      "prop_refuge",
      background = styleColorBar(range(refuge_table$prop_refuge), "lightblue"),
      backgroundSize = "100% 90%",
      backgroundRepeat = "no-repeat",
      backgroundPosition = "center"
    )
} else {
  knitr::kable(refuge_table, digits = 2)
}

```

## Refuge plot
```{r}
#| label: fig-refuge-plot
#| include: true
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Kōura refuge use over the course of the experiment, showing the proportion of time spent in refuge across three periods (before, during, and after predator exposure) for each treatment (catfish: dark grey solid line; eel: medium grey dashed line; control: light grey dot-dash line) under light (top) and dark (bottom) conditions. Black dashed vertical lines indicate the start and end of predator exposure and grey vertical lines show individual observation times."

# calculate the proportions of location for each treatment for each time step
prop_location_time <- koura_long_ref %>%
  filter(!is.na(location), location != "") %>%
  count(treatment, time, light, location = location, name = "n") %>%
  group_by(treatment, time, light) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  mutate(treatment = factor(treatment, levels = c("catfish","eel","control")),
         location  = factor(location,  levels = c("refuge","open")),
         light     = factor(light,     levels = c("light", "dark")))

refuge_plot <- prop_location_time %>%
  dplyr::filter(location == "refuge") %>%
  ggplot2::ggplot(aes(time, prop, colour = treatment, group = treatment, linetype = treatment)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = c(60, 240), linetype = "dashed") +
  geom_vline(xintercept = c(0, 5, 10, 20, 30, 45, 60, 61, 65, 70, 80, 90, 120, 150, 210, 240, 245, 260, 275, 290, 300), alpha = .15) +
  annotate("text", x = 30,  y = 1, label = "Before", size = 3) +
  annotate("text", x = 150, y = 1, label = "During", size = 3) +
  annotate("text", x = 270, y = 1, label = "After",  size = 3) +
  scale_colour_manual(values = c("catfish" = "grey20",  "eel"  = "grey40",  "control" = "grey60"),
                      labels = c("catfish" = "Catfish", "eel" = "Eel", "control" = "Control")) +
  scale_linetype_manual(values = c("solid", "dashed", "dotdash"),
                        labels = c("catfish" = "Catfish", "eel" = "Eel", "control" = "Control")) +
  facet_grid(light ~ ., scales = "free_x", labeller = as_labeller(str_to_title)) +
  labs(x = "Time (minutes)", y = "Proportion of refuge use", colour = "Treatment", linetype = "Treatment")

ggsave(file.path(out_dir, "fig-refuge-plot.png"), refuge_plot, width = 8, height = 5, dpi = 300)

refuge_plot

```

## Refuge stats
```{r}
#| label: refuge-stats
#| include: false

koura_stat <- koura_long_ref %>%
  select(treatment, exp_id, round, tank, time, period, light, location, koura_id, size_mm1, weight_g1) %>%
  mutate(
    location = factor(location, levels = c("refuge", "open")),
    period   = factor(period,   levels = c("before", "during", "after")),
    tank     = factor(tank))

round_summary <- koura_stat %>%
  group_by(treatment, exp_id, round, tank, light, period, koura_id, size_mm1, weight_g1) %>%
  summarise(n_total  = n(),n_refuge = sum(location == "refuge"), .groups  = "drop") %>%
  mutate(treatment = relevel(factor(treatment), ref = "control"),
    light = relevel(factor(light), ref = "light"),   
    period = relevel(factor(period), ref = "before"))

# Light condition only (rounds 1-4)
light_data <- round_summary %>% filter(light == "light")
m_light <- glmmTMB(cbind(n_refuge, n_total - n_refuge) ~ treatment + period + (1|tank) + (1|round), 
                    data = light_data, family = betabinomial)

# Dark condition only (rounds 5-8)
dark_data <- round_summary %>% filter(light == "dark")
m_dark <- glmmTMB(cbind(n_refuge, n_total - n_refuge) ~ treatment + period + (1|tank) + (1|round), 
                   data = dark_data, family = betabinomial)

summary(m_light)
summary(m_dark)

# Model diagnostics 
sim_light <- simulateResiduals(m_light)
plot(sim_light)
testDispersion(sim_light)
testZeroInflation(sim_light)

# Model diagnostics 
sim_dark <- simulateResiduals(m_dark)
plot(sim_dark)
testDispersion(sim_dark)
testZeroInflation(sim_dark)

## Fit light model WITHOUT round RE to put light in there
m_light_effect <- glmmTMB(cbind(n_refuge, n_total - n_refuge) ~ treatment + light + period + (1|tank), data = round_summary, family = betabinomial)

summary(m_light_effect)

```

## Fixed effects - Light and Dark
```{r}
#| label: tbl-fixed-effects-stratified
#| include: true
#| tbl-cap: "Fixed effects of beta-binomial GLMMs for kōura refuge use in light and dark conditions."

# Light condition
fixed_light <- as.data.frame(summary(m_light)$coefficients$cond) %>%
  rownames_to_column("term") %>%
  mutate(condition = "Light", across(where(is.numeric), round, 3))

# Dark condition
fixed_dark <- as.data.frame(summary(m_dark)$coefficients$cond) %>%
  rownames_to_column("term") %>%
  mutate(condition = "Dark", across(where(is.numeric), round, 3))

# Combine
fixed_bb_stratified <- bind_rows(fixed_light, fixed_dark) %>%
  select(condition, everything())

write.csv(fixed_bb_stratified, file.path(out_dir, "tbl-fixed-effects-stratified.csv"), row.names = FALSE)

knitr::kable(fixed_bb_stratified, col.names = c("Condition", "Term", "Estimate", "Std. Error", "z value", "p value"))
```

## Emmeans - Light and Dark
```{r}
#| label: tbl-emmeans-stratified
#| include: true
#| tbl-cap: "Pairwise contrasts for kōura refuge use in light and dark conditions. Results are shown separately for each light condition from beta-binomial GLMMs. Odds ratios and p-values are shown for treatment and period comparisons. P-values are Sidak-adjusted for multiple comparisons within each factor."

# Light
emm_treat_light <- pairs(emmeans(m_light, ~ treatment, type = "response"), adjust = "tukey") %>%
  as.data.frame() %>% mutate(condition = "Light")

emm_period_light <- pairs(emmeans(m_light, ~ period, type = "response"), adjust = "tukey") %>%
  as.data.frame() %>% mutate(condition = "Light")

# Dark
emm_treat_dark <- pairs(emmeans(m_dark, ~ treatment, type = "response"), adjust = "tukey") %>%
  as.data.frame() %>% mutate(condition = "Dark")

emm_period_dark <- pairs(emmeans(m_dark, ~ period, type = "response"), adjust = "tukey") %>%
  as.data.frame() %>% mutate(condition = "Dark")

# Combine
emmeans_stratified <- bind_rows(emm_treat_light, emm_period_light, emm_treat_dark, emm_period_dark) %>%
  select(condition, everything()) %>%
  mutate(across(where(is.numeric), round, 3))

write.csv(emmeans_stratified, file.path(out_dir, "tbl-emmeans-stratified.csv"), row.names = FALSE)

knitr::kable(emmeans_stratified, col.names = c("Condition", "Contrast", "Odds ratio", "SE", "df", "Null", "z value", "p value"))
```

## Emmeans plots - Light and Dark
```{r}
#| label: fig-emmeans-stratified
#| include: true
#| fig-width: 7
#| fig-height: 3.5
#| fig-cap: "Predicted marginal probabilities of refuge use by treatment and experimental period in light and dark conditions. Results are shown separately for light (top) and dark (bottom) conditions, each fitted with a beta-binomial GLMM. Points represent model-estimated marginal means with 95% confidence intervals shown as error bars. Different letters (a, b) indicate statistically significant differences based on Tukey-adjusted pairwise comparisons within each factor. The dashed line at 0.5 marks the threshold where kōura spend half their time in refuge."

# Get predictions AND compute CLDs for both models
# Light condition
emm_treat_light_pred <- as.data.frame(emmeans(m_light, ~ treatment, type = "response")) %>%
  mutate(condition = "Light", factor = "Treatment", x = str_to_title(treatment))
cld_treat_light <- as.data.frame(multcomp::cld(emmeans(m_light, ~ treatment, type = "response"), 
                                                Letters = letters, adjust = "sidak")) %>%
  mutate(condition = "Light", factor = "Treatment", x = str_to_title(treatment))

emm_period_light_pred <- as.data.frame(emmeans(m_light, ~ period, type = "response")) %>%
  mutate(condition = "Light", factor = "Period", x = str_to_title(period))
cld_period_light <- as.data.frame(multcomp::cld(emmeans(m_light, ~ period, type = "response"), 
                                                 Letters = letters, adjust = "sidak")) %>%
  mutate(condition = "Light", factor = "Period", x = str_to_title(period))

# Dark condition
emm_treat_dark_pred <- as.data.frame(emmeans(m_dark, ~ treatment, type = "response")) %>%
  mutate(condition = "Dark", factor = "Treatment", x = str_to_title(treatment))
cld_treat_dark <- as.data.frame(multcomp::cld(emmeans(m_dark, ~ treatment, type = "response"), 
                                               Letters = letters, adjust = "sidak")) %>%
  mutate(condition = "Dark", factor = "Treatment", x = str_to_title(treatment))

emm_period_dark_pred <- as.data.frame(emmeans(m_dark, ~ period, type = "response")) %>%
  mutate(condition = "Dark", factor = "Period", x = str_to_title(period))
cld_period_dark <- as.data.frame(multcomp::cld(emmeans(m_dark, ~ period, type = "response"), 
                                                Letters = letters, adjust = "sidak")) %>%
  mutate(condition = "Dark", factor = "Period", x = str_to_title(period))

# Combine all (Period first, Treatment second for left-to-right order)
emm_all_stratified <- bind_rows(emm_period_light_pred, emm_treat_light_pred,
                                emm_period_dark_pred, emm_treat_dark_pred) %>%
  mutate(condition = factor(condition, levels = c("Light","Dark")),
  factor = factor(factor, levels = c("Treatment", "Period")),
  x = factor(x, levels = c("Catfish", "Eel", "Control", "Before", "During", "After")))


cld_all_stratified <- bind_rows(cld_period_light, cld_treat_light,
                                cld_period_dark, cld_treat_dark) %>%
  mutate(condition = factor(condition, levels = c("Light","Dark")),
  factor = factor(factor, levels = c("Treatment", "Period")),
  x = factor(x, levels = c("Catfish", "Eel", "Control", "Before", "During", "After")))


# Plot with letters
emmeans_plot_stratified <- ggplot(emm_all_stratified, aes(x = x, y = prob, ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  geom_text(data = cld_all_stratified, aes(x = x, y = asymp.UCL + 0.08, label = .group), inherit.aes = FALSE, size = 4) +
  facet_grid(condition ~ factor, scales = "free_x") +
  scale_y_continuous("Predicted probability of refuge use", limits = c(0, 1)) +
  scale_x_discrete(NULL) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") 

ggsave(file.path(out_dir, "fig-emmeans-stratified.png"), emmeans_plot_stratified, width = 5, height = 4, dpi = 300)

emmeans_plot_stratified
```

```{r}
#| label: fig-emmeans-treatment-plot
#| include: true
#| fig-width: 3
#| fig-height: 3

emm_treat_light_pred <- as.data.frame(emmeans(m_light, ~ treatment, type = "response")) %>%
  mutate(condition = "Light", x = str_to_title(treatment))
cld_treat_light <- as.data.frame(multcomp::cld(emmeans(m_light, ~ treatment, type = "response"), 
                                                Letters = letters, adjust = "sidak")) %>%
  mutate(condition = "Light", x = str_to_title(treatment))

emm_treat_dark_pred <- as.data.frame(emmeans(m_dark, ~ treatment, type = "response")) %>%
  mutate(condition = "Dark", x = str_to_title(treatment))
cld_treat_dark <- as.data.frame(multcomp::cld(emmeans(m_dark, ~ treatment, type = "response"), 
                                               Letters = letters, adjust = "sidak")) %>%
  mutate(condition = "Dark", x = str_to_title(treatment))

emm_treat_all <- bind_rows(emm_treat_light_pred, emm_treat_dark_pred) %>%
  mutate(condition = factor(condition, levels = c("Dark", "Light")),
         x = factor(x, levels = c("Catfish", "Eel", "Control")))

cld_treat_all <- bind_rows(cld_treat_light, cld_treat_dark) %>%
  mutate(condition = factor(condition, levels = c("Dark", "Light")),
         x = factor(x, levels = c("Catfish", "Eel", "Control")))

treat_plot <- ggplot(emm_treat_all, aes(x = x, y = prob, ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.2) +
  geom_text(data = cld_treat_all, aes(x = x, y = asymp.UCL + 0.08, label = .group), 
            inherit.aes = FALSE, size = 4) +
  facet_wrap(~ condition) +
  scale_y_continuous("Predicted Probability of Refuge Use", limits = c(0, 1)) +
  scale_x_discrete(NULL) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50")

ggsave(file.path(out_dir, "fig-treatment-emmeans.png"), treat_plot, width = 4, height = 3, dpi = 300)

treat_plot

```

## Summary table of koura used
```{r}
#| label: tbl-koura
#| include: false
#| tbl-cap: "Summary of kōura used in experiments"

grps <- list(c("treatment","light"), "treatment", "light", character(0))
tags <- c("treatment+light","treatment","light","overall")

summary_animals <- Map(function(g, tag)
  df_NCE_raw %>%
    group_by(across(all_of(g))) %>%
    summarise(n_trials = n(), mean_size = mean(size_mm1, na.rm=TRUE), sd_size = sd(size_mm1, na.rm=TRUE), min_size  = min(size_mm1, na.rm = TRUE), max_size = max(size_mm1, na.rm = TRUE), mean_weight = mean(weight_g1, na.rm=TRUE), sd_weight = sd(weight_g1,       na.rm=TRUE),min_weight = min(weight_g1, na.rm = TRUE),
      max_weight = max(weight_g1, na.rm = TRUE),.groups="drop") %>%
    mutate(grouping=tag) %>%
    { if(!"treatment"%in%g) mutate(., treatment="All") else . } %>%
    { if(!"light"%in%g)     mutate(., light="All")     else . },
  grps, tags) %>%
  bind_rows() %>%
  select(grouping, treatment, light, everything()) %>%
  arrange(factor(grouping, levels=tags), treatment, light)

# need to make into nice table
knitr::kable(summary_animals, digits = 2)

```

## Plots individual koura
```{r}
#| label: fig-koura
#| include: false
#| fig-width: 5
#| fig-height: 5
#| fig-cap: "your caption"


loc_order <- c("ref_back","ref_open","ref_next","wall","floor","barrier")
beh_order <- c("static","walk","swim","climb")

koura_location_time <- koura_long %>%
  mutate(location = factor(location, levels = loc_order), treatment = factor(treatment, levels = c("catfish", "eel", "control")))


koura_behaviour_time <- koura_long %>%
  mutate(behaviour = factor(behaviour, levels = beh_order), treatment = factor(treatment, levels = c("catfish", "eel", "control")))

# --- Plots ---
p_loc_time <- ggplot(koura_long_ref, aes(time, location, col = treatment)) +
  geom_point() +  
  geom_line(aes(group = koura_id)) +
  geom_vline(xintercept = c(60, 240), linetype = "dashed") +
  facet_wrap(koura_id ~ light) +
  labs(x = "Time (min)", y = "Location")+
  theme(legend.position = "none")

p_beh_time <- ggplot(koura_behaviour_time, aes(time, behaviour, col = treatment)) +
  geom_point() +
  geom_line(aes(group = koura_id))+
  geom_vline(xintercept = c(60, 240), linetype = "dashed") +
  facet_wrap(koura_id ~ light) +
  labs(x = "Time (min)", y = "Behaviour")

# --- Combine ---
koura_time_plot <- p_loc_time + p_beh_time
koura_time_plot

```

## Koura length refuge use
```{r}
#| label: koura-refuge-use
#| include: false

# Calculate proportion of time each koura spent in refuge (overall)
refuge_by_koura <- koura_long_ref %>%
  group_by(exp_id , treatment, size_mm1) %>%
  summarise(
    total_obs = n(),
    refuge_obs = sum(location == "refuge", na.rm = TRUE),
    prop_refuge = refuge_obs / total_obs,
    .groups = "drop"
  )

# Check what you have
print(refuge_by_koura)

# Spearman correlation - separate by treatment
cor_catfish <- refuge_by_koura %>% filter(treatment == "catfish")
cor_eel     <- refuge_by_koura %>% filter(treatment == "eel")
cor_control <- refuge_by_koura %>% filter(treatment == "control")

# Run Spearman tests
spearman_catfish <- cor.test(cor_catfish$size_mm1, cor_catfish$prop_refuge, 
                              method = "spearman")
spearman_eel     <- cor.test(cor_eel$size_mm1,     cor_eel$prop_refuge,     
                              method = "spearman")
spearman_control <- cor.test(cor_control$size_mm1, cor_control$prop_refuge, 
                              method = "spearman")

# Print results
spearman_catfish
spearman_eel
spearman_control

p_raw <- c(spearman_catfish$p.value, spearman_eel$p.value, spearman_control$p.value)
p.adjust(p_raw, method = "bonferroni")

# Extract formatted p-values and correlations
rho_catfish <- round(spearman_catfish$estimate, 3)
rho_eel <- round(spearman_eel$estimate, 3)
rho_control <- round(spearman_control$estimate, 3)
p_adj <- p.adjust(p_raw, method = "bonferroni")

# Optional quick scatter plots to visualise
par(mfrow = c(1, 3))

plot(cor_catfish$size_mm1, cor_catfish$prop_refuge,
     main = "Catfish", xlab = "Size (mm)", ylab = "Proportion time in refuge",
     pch = 19)
abline(lm(prop_refuge ~ size_mm1, data = cor_catfish), col = "red")

plot(cor_eel$size_mm1, cor_eel$prop_refuge,
     main = "Eel", xlab = "Size (mm)", ylab = "Proportion time in refuge",
     pch = 19)
abline(lm(prop_refuge ~ size_mm1, data = cor_eel), col = "green")

plot(cor_control$size_mm1, cor_control$prop_refuge,
     main = "Control", xlab = "Size (mm)", ylab = "Proportion time in refuge",
     pch = 19)
abline(lm(prop_refuge ~ size_mm1, data = cor_control), col = "blue")

par(mfrow = c(1, 1))

```


## Koura before behaviour
```{r}
#| label: koura-before-behaviour
#| include: false

df_before <- koura_long_ref %>% 
  filter(period =="before") %>%
  filter(!is.na(location), location != "") %>%
  count(exp_id, treatment, light, location = location, name = "n") %>%
  group_by(exp_id) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  mutate(treatment = factor(treatment, levels = c("catfish","eel","control")),
    location  = factor(location,  levels = c("refuge","open"))) %>%
  filter(location == "refuge")

# Coefficient of Variation (CV)
before_cv_ref <- df_before %>%
  group_by(light) %>%
  summarise(cv = sd(prop) / mean(prop) * 100)

#leveneTest(prop ~ light, data = df_before)
#fligner.test(prop ~ light, data = df_before)


time_compare <- koura_long_ref %>%
  filter(time == 5 | time == 60) %>%
  filter(!is.na(location), location != "") %>%
  group_by(time, location) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = location, values_from = n, values_fill = 0)

# contingency table
cont_table <- time_compare %>%
  select(refuge, open) %>%
  as.matrix()

rownames(cont_table) <- c("time_5", "time_60")
```



# Feeding experiment
## Feeding table
```{r}
#| label: tbl-feeding
#| include: true
#| tbl-cap: "Mean feeding rate per gram body weight by treatment and starvation day."

df_feeding <- df_Feeding_raw %>%
  filter(sex != "F") %>%
  mutate(pellets_eaten_24h = total_fed - left_after_24h,
    pellets_per_g = pellets_eaten_24h / weight_g,
    koura_id = as.factor(koura_id),
    treatment = as.factor(treatment),
    day_starve = as.factor(day_starve),
    treatment = factor(treatment, levels = c("catfish", "eel", "control")))


# summary table
feeding_summary <- df_feeding %>%
  group_by(treatment, day_starve) %>%
  summarise(
    mean = round(mean(pellets_per_g, na.rm = TRUE), 3),
    se   = round(sd(pellets_per_g, na.rm = TRUE) / sqrt(n()), 3),
    .groups = "drop"
  )

# helper for inline use
get_mean_se <- function(trt, starve) {
  row <- feeding_summary %>% filter(treatment == trt, day_starve == starve)
  paste0(row$mean, " ± ", row$se)
}

knitr::kable(feeding_summary, col.names = c("Treatment", "Days starved", "Mean (pellets/g)", "SE"))
```

## Feeding plot
```{r}
#| label: fig-feeding-plot
#| include: true
#| fig-width: 5
#| fig-height: 4
#| fig-cap: "Mean (±SE) kōura feeding rate (pellets consumed per gram wet weight kōura) across predator treatments (catfish, eel, control) and starvation periods (0, 3, and 4 days)."

feeding_plot <- ggplot(df_feeding, aes(x = factor(day_starve),  y = pellets_per_g, fill = treatment)) +
  stat_summary(fun = mean, geom = "bar", position = position_dodge(0.9), colour = "black", linewidth = 0.4, pattern_fill = "black",pattern_colour = "black", pattern_density = 0.3, pattern_spacing = 0.05) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.25, linewidth = 0.5,  position = position_dodge(0.9)) +
  scale_fill_manual(values = c("catfish" = "gray20",  "eel"  = "grey40",  "control" = "grey60"),
                   labels = c("catfish" = "Catfish", "eel" = "Eel", "control" = "Control")) +
  labs(x = "Starvation period (days)",  y = "Feeding rate (pellets consumed g WW⁻¹)",  fill = "Treatment") +
  theme(legend.position = "right")

ggsave(file.path(out_dir, "fig-feeding-plot.png"), feeding_plot, width = 5, height = 4, dpi = 300)

feeding_plot

```

## Feeding stats
```{r}
#| label: feeding-stats
#| include: false

# Model selection on LMER models
feeding_models_lmer <- list(
  m0 = lmer(pellets_per_g ~ 1 + (1 | exp_tank), data = df_feeding),
  m1 = lmer(pellets_per_g ~ treatment * day_starve + (1 | exp_tank), data = df_feeding),
  m2 = lmer(pellets_per_g ~ treatment + day_starve + (1 | exp_tank), data = df_feeding),
  m5 = lmer(pellets_per_g ~ treatment + (1 | exp_tank), data = df_feeding),
  m6 = lmer(pellets_per_g ~ day_starve + (1 | exp_tank), data = df_feeding))

comparison_lmer <- data.frame(
  model  = names(feeding_models_lmer),
  AIC    = sapply(feeding_models_lmer, AIC),
  BIC    = sapply(feeding_models_lmer, BIC),
  logLik = sapply(feeding_models_lmer, \(m) as.numeric(logLik(m))),
  AICc   = sapply(feeding_models_lmer, AICc),
  k      = sapply(feeding_models_lmer, \(m) attr(logLik(m), "df"))) %>%
  arrange(AICc) %>%
  mutate(delta_AICc = round(AICc - min(AICc), 2))

print(comparison_lmer)

# Best model by AICc: m0 (intercept-only, delta_AICc = 0.00)
# However, m6 (day_starve only, delta_AICc = 0.86) is retained as it is 
# biologically interpretable and accounts for tank-level variation while 
# testing starvation effects. Treatment does not improve model fit 
# (m2 delta_AICc = 14.38, m1 delta_AICc = 32.07).

# Final model with tank as random effect
lmm_feed <- lmer(pellets_per_g ~ day_starve + (1 | exp_tank), data = df_feeding)
summary(lmm_feed)

# Extract p-values and statistics for inline reporting
# From summary output (t-tests, not Chisq)
summary_lmm <- summary(lmm_feed)
p_day_starve3 <- round(summary_lmm$coefficients["day_starve3", "Pr(>|t|)"], 4)
p_day_starve4 <- round(summary_lmm$coefficients["day_starve4", "Pr(>|t|)"], 4)
p_shapiro_feed <- round(shapiro.test(residuals(lmm_feed))$p.value, 3)

# emmeans for inline values
emm_feed_df <- as.data.frame(emmeans(lmm_feed, ~ day_starve)) %>%
  mutate(across(where(is.numeric), round, 3))

# Model diagnostics for lmer
par(mfrow = c(2, 2))
plot(lmm_feed)
par(mfrow = c(1, 1))

# Check normality of residuals
shapiro.test(residuals(lmm_feed))

# Pairwise comparisons for day_starve (Tukey-adjusted)
emm_day_starve <- emmeans(lmm_feed, ~ day_starve)
pairs(emm_day_starve, adjust = "tukey")

# Stratified analysis by starvation period (to avoid confounding)
# Use lm here since each stratum has limited observations
df_day0 <- df_feeding %>% filter(day_starve == 0)
df_day3 <- df_feeding %>% filter(day_starve == 3)
df_day4 <- df_feeding %>% filter(day_starve == 4)

lm_day0 <- lm(pellets_per_g ~ treatment, data = df_day0)
lm_day3 <- lm(pellets_per_g ~ treatment, data = df_day3)
lm_day4 <- lm(pellets_per_g ~ treatment, data = df_day4)

summary(lm_day0)
summary(lm_day3)
summary(lm_day4)

# Pairwise comparisons for treatment within each starvation period
pairs(emmeans(lm_day0, ~ treatment), adjust = "tukey")
pairs(emmeans(lm_day3, ~ treatment), adjust = "tukey")
pairs(emmeans(lm_day4, ~ treatment), adjust = "tukey")

# Check normality of residuals for each
shapiro.test(residuals(lm_day0))
shapiro.test(residuals(lm_day3))
shapiro.test(residuals(lm_day4))

kruskal.test(pellets_per_g ~ treatment, data = df_day0)


# Extract t and p values from main model
p_day_starve3 <- round(summary_lmm$coefficients["day_starve3", "Pr(>|t|)"], 3)
p_day_starve4 <- round(summary_lmm$coefficients["day_starve4", "Pr(>|t|)"], 3)

# Shapiro-Wilk for day 0
sw_day0 <- shapiro.test(residuals(lm_day0))
sw_day0_p <- round(sw_day0$p.value, 4)

# Kruskal-Wallis result for day 0
kw_day0 <- kruskal.test(pellets_per_g ~ treatment, data = df_day0)
kw_day0_chi <- round(kw_day0$statistic, 4)
kw_day0_p <- round(kw_day0$p.value, 4)

# p-values from stratified analyses
p_day3_treat <- round(anova(lm_day3)[1, "Pr(>F)"], 4)
p_day4_treat <- round(anova(lm_day4)[1, "Pr(>F)"], 4)

# Get means from day 4 stratified model
emm_day4 <- as.data.frame(emmeans(lm_day4, ~ treatment))

get_feed_mean_se <- function(treatment_name) {
  row <- emm_day4 %>% filter(treatment == treatment_name)
  paste0(round(row$emmean, 3), " ± ", round(row$SE, 3))
}

```

### feeding anova tbl
```{r}
#| label: tbl-feeding-summary
#| include: true
#| tbl-cap: "Fixed effects from the linear mixed model of kōura feeding rate (pellets consumed per gram body weight). Starvation period included as fixed effect with tank as random intercept."

# Extract summary table from lmer
summary_lmm <- summary(lmm_feed)
feed_summary <- as.data.frame(summary_lmm$coefficients) %>%
  rownames_to_column("term") %>%
  mutate(across(where(is.numeric), round, 3))

write.csv(feed_summary, file.path(out_dir, "tbl-feeding-summary.csv"), row.names = FALSE)

knitr::kable(feed_summary, 
             col.names = c("Term", "Estimate", "Std. Error", "df", "t value", "p value"),
             digits = 3)
```

# Session info
```{r}
#| label: session-info
#| include: true
#| code-fold: true
sessionInfo()
```